manuscript_outline

### 2025-07-22 docs/manuscript/paper.Rmd
add bcl's edits

# A generalizable tool for predicting developmental phenology for wild poikilotherms.

@@

keywords: developmental phenology; hatchR; local adaptation,;R, reaction norm; Shiny

@@

fmt <- "~R^2 ==  %.4f"
lab1 <- sprintf(fmt, ascaphus_mod$r_squared)
p1 <- ascaphus_mod$pred_plot +
  labs(x = "Incubation Temperature (°C)", y = "Days to Hatch") +
  lims(y = c(0, 50)) +
  annotate("text", x = 10, y = c(35), label = c(lab1),  hjust = 0, parse = TRUE) +
  theme_classic(base_size = 16)

@@

p2.1.a <- ggplot() +
  geom_point(data = temperature_data_cold, aes(x = dates, y = temperatures)) +
  geom_line(data = temperature_data_cold, aes(x = dates, y = temperatures)) +
  geom_rect(data = froggy_hatch_cold, aes(xmin = start, xmax = stop, ymax = index - .15, ymin = index - .5, fill = species)) +
  geom_label(data = froggy_hatch_cold, aes(x = start + (stop - start) / 1.45, y = (index - 0.325), label = days_to_develop)) +
  labs(x = "Date", y = "Temperature (°C)") +
  scale_fill_manual(values = c("darkolivegreen4", "deepskyblue4", "grey23", "purple4"), name = "Species") +
  theme_classic(base_size = 16) +
  theme(legend.position = "none")

@@

p2.1.b <- ggplot() +
  geom_point(data = temperature_data_warm, aes(x = dates, y = temperatures)) +
  geom_line(data = temperature_data_warm, aes(x = dates, y = temperatures)) +
  geom_rect(data = froggy_hatch_warm, aes(xmin = start, xmax = stop, ymax = index - .15, ymin = index - .5, fill = species)) +
  geom_label(data = froggy_hatch_warm, aes(x = start + (stop - start) / 1.45, y = (index - 0.325), label = days_to_develop)) +
  labs(x = "Date", y = "Temperature (°C)") +
  scale_fill_manual(values = c("darkolivegreen4", "deepskyblue4", "grey23", "purple4"), name = "Species") +
  theme_classic(base_size = 16) +
  theme(legend.position = "none")

@@

p2.2 <- ef_vals |>
  tibble() |>
  ggplot(aes(x = temperature, y = effective_value, color = frog_species)) +
  geom_line() +
  geom_point() +
  geom_abline(intercept = 0, slope = .01, linetype = "dashed") +
  labs(x = "Daily Average Temperature (°C)", y = "Effective Value") +
  scale_color_manual(values = c("darkolivegreen4", "deepskyblue4", "grey23", "purple4"), name = "Species") +
  theme_classic(base_size = 16) +
  theme(legend.position = c(0.2,0.8))

@@

fig4 <- ef_vals_pops |>
  tibble() |>
  ggplot(aes(x = temperature, y = effective_value, color = beetle_pops)) +
  geom_line() +
  geom_point() +
  # geom_abline(intercept = 0, slope = .01, linetype = "dashed") +
  labs(x = "Daily Average Temperature (°C)", y = "Effective Value", title = "Figure 4") +
  scale_color_brewer(palette = "Dark2", name = "Beetle Populations") +
  theme_classic(base_size = 16) +
  theme(legend.position = c(0.25, 0.75))

@@

| Class | Order | Genus | Species | Study |
|:--------------|:--------------|:--------------|:--------------|:--------------|
| Amphibia | Anura | *Lithobates* | *L. sylvaticus, L. pipiens, L. clamitans, L. palustris* | @moore1939 |
|  |  | *Ascaphus* | *A. truei* | @brown1975 |
|  | Urodela | *Ambystoma* | *A. gracile* | @brown1976 |
| Reptilia | Squamata | *Sceloporus* | *S. undulatus* | @angilletta2000 |
|  |  | *Podarcis* | *P. muralis* | @vandamme1992 |
|  | Testudines | *Mauremys* | *M. reevesii* | @du2007 |
|  |  |  | 181 species | 141 studies in @while2018 |
| Insecta | Plecoptera | *Nemurella* | *N. pictetii* | @brittain1978, @elliott1984 |
|  |  | *Capnia* | *C. atra* | @brittain1984 |
|  |  | *Capnia* | *C. bifrons* | @elliott1986 |
|  |  | *Mesocapnia* | *M. oenone* | @brittain1984 |
|  |  | *Taeniopteryx* | *T. nebulosa* | @brittain1977 |
|  | Coleoptera | *Colaphellus* | *C. bowringi* | @tang2017 |
|  | Odonata | 7 genera | 14 species | Developmental equations in @pritchard1987 |
|  | Culicidae | *Aedes* | *A. sticticus, A. vexans* |  |
|  |  | *Anopheles* | *A. quadrimaculatus* |  |
|  |  | *Toxorhynchites* | *T. brevipalpis* |  |
| Malacostraca | Decapoda | *Pontastacus* | *P. leptodactylus* | @aydin2004 |
| Copepoda | Calanoida | 10 genera | 28 species | @forster2011 |
|  | Cyclopoida | *Limnoithona* | *L. tetraspina* |  |
|  |  | *Oithona* | *O. davisae* |  |
|  | Harpacticoida | *Microsetella* | *M. norvegica* |  |
|  |  | *Mesochra* | *M.lilljeborgi* |  |
|  | Poecilostomatoida | *Oncaea* | *O. venusta* |  |
| Cephalopoda | Octopoda | *Octopus* | *O. vulgaris, O. mimus* | @márquez2021 |
|  | Myopsida | *Loligo* | *L. vulgaris, L. reynaudii* |  |
|  | Oegopsida | *Illex* | *I. coindetii, I. illecebrosus* |  |
|  |  | *Todarodes* | *T.pacificus* |  |
|  |  | *Ommastrephes* | *O. bartramii* |  |
| Asteroidea | Valvatida | *Odontaster* | *O. meridionalis, O. validus* | @hoegh-guldberg1995 |
|  |  | *Asterina* | *A. miniata* |  |
|  |  | *Acanthaster* | *A. planci* |  |

@@

Figure 1: Custom hatching phenology model for coastal tailed frogs (*Ascaphus truei*). Panel A represents the model fit and raw data used to generate the effective value model and panel B are the effective values for daily temperatures between 6 and 20 °C. A dashed line with a 0.01 increase for every degree increase in included for reference.

Figure 2: Modeled changes in developmental timing for coastal tailed frogs (*Ascaphus truei*) from 1991 to 2020. Panel A: Marginal mean trend across all sites showing a significant decline in days to hatching over time. Panel B: Site-specific (COMID) trends in predicted days to hatch, colored by mountain range. Estimates are based on the best fitting model, with stream temperatures from @siegel2023 and developmental timing modeled using hatchR. Shaded ribbons represent 95% confidence intervals.

>>>>

# A generalizable tool for predicting developmental phenology for wild poikilotherms

@@

keywords: developmental phenology; hatchR; local adaptation; R; reaction norm; Shiny

@@

fmt <- "~R^2 ==  %.4f"
lab1 <- sprintf(fmt, ascaphus_mod$r_squared)
p1 <- ascaphus_mod$pred_plot +
  labs(x = "Incubation Temperature (°C)", y = "Days to Hatch") +
  lims(y = c(0, 50), x =c(5,20)) +
  annotate("text", x = 10, y = c(35), label = c(lab1),  hjust = 0, parse = TRUE) +
  theme_classic(base_size = 16)

@@

p2.1.a <- ggplot() +
  geom_point(data = temperature_data_cold, aes(x = dates, y = temperatures)) +
  geom_line(data = temperature_data_cold, aes(x = dates, y = temperatures)) +
  geom_rect(data = froggy_hatch_cold, aes(xmin = start, xmax = stop, ymax = index - .15, ymin = index - .5, fill = species)) +
  geom_label(data = froggy_hatch_cold, aes(x = start + (stop - start) / 1.45, y = (index - 0.325), label = days_to_develop)) +
  labs(x = "Date", y = "Temperature (°C)") +
  scale_fill_manual(values = c("darkolivegreen4", "deepskyblue4", "grey23", "purple4"), name = "Species") +
  theme_classic(base_size = 14) +
  theme(legend.position = "none")

@@

p2.1.b <- ggplot() +
  geom_point(data = temperature_data_warm, aes(x = dates, y = temperatures)) +
  geom_line(data = temperature_data_warm, aes(x = dates, y = temperatures)) +
  geom_rect(data = froggy_hatch_warm, aes(xmin = start, xmax = stop, ymax = index - .15, ymin = index - .5, fill = species)) +
  geom_label(data = froggy_hatch_warm, aes(x = start + (stop - start) / 1.45, y = (index - 0.325), label = days_to_develop)) +
  labs(x = "Date", y = "Temperature (°C)") +
  scale_fill_manual(values = c("darkolivegreen4", "deepskyblue4", "grey23", "purple4"), name = "Species") +
  theme_classic(base_size = 14) +
  theme(legend.position = "none")

@@

p2.2 <- ef_vals |>
  tibble() |>
  ggplot(aes(x = temperature, y = effective_value, color = frog_species)) +
  geom_line() +
  geom_point() +
  geom_abline(intercept = 0, slope = .01, linetype = "dashed") +
  labs(x = "Daily Average Temperature (°C)", y = "Effective Value") +
  scale_color_manual(values = c("darkolivegreen4", "deepskyblue4", "grey23", "purple4"), name = "Species") +
  theme_classic(base_size = 14) +
  theme(legend.position = c(0.2,0.8))

@@

fig4 <- ef_vals_pops |>
  tibble() |>
  ggplot(aes(x = temperature, y = effective_value, color = beetle_pops)) +
  geom_line() +
  geom_point() +
  # geom_abline(intercept = 0, slope = .01, linetype = "dashed") +
  labs(x = "Daily Average Temperature (°C)", y = "Effective Value") +
  scale_color_brewer(palette = "Dark2", name = "Beetle Populations") +
  theme_classic(base_size = 16) +
  theme(legend.position = c(0.25, 0.75))

fig4 <- fig4 + plot_annotation(title = "Figure 4")

@@

| Class | Order | Genus | Species | Study |
|:--------------|:--------------|:--------------|:--------------|:--------------|
| Amphibia | Anura | *Ascaphus* | *A. truei* | @brown1975 |
|  |  | *Lithobates* | *L. clamitans, L. palustris, L. pipiens, L. sylvaticus* | @moore1939 |
|  | Urodela | *Ambystoma* | *A. gracile* | @brown1976 |
| Asteroidea | Valvatida | *Acanthaster* | *A. planci* | @hoegh-guldberg1995 |
|  |  | *Asterina* | *A. miniata* |  |
|  |  | *Odontaster* | *O. meridionalis, O. validus* |  |
| Cephalopoda | Myopsida | *Loligo* | *L. reynaudii, L. vulgaris* | @márquez2021 |
|  | Octopoda | *Octopus* | *O. mimus, O. vulgaris* |  |
|  | Oegopsida | *Illex* | *I. coindetii, I. illecebrosus* |  |
|  |  | *Ommastrephes* | *O. bartramii* |  |
|  |  | *Todarodes* | *T. pacificus* |  |
| Copepoda | Calanoida | 10 genera | 28 species | @forster2011 |
|  | Cyclopoida | *Limnoithona* | *L. tetraspina* |  |
|  |  | *Oithona* | *O. davisae* |  |
|  | Harpacticoida | *Mesochra* | *M. lilljeborgi* |  |
|  |  | *Microsetella* | *M. norvegica* |  |
|  | Poecilostomatoida | *Oncaea* | *O. venusta* |  |
| Insecta | Coleoptera | *Colaphellus* | *C. bowringi* | @tang2017 |
|  | Culicidae | *Aedes* | *A. sticticus, A. vexans* | Developmental equations in @pritchard1987 |
|  |  | *Anopheles* | *A. quadrimaculatus* |  |
|  |  | *Toxorhynchites* | *T. brevipalpis* |  |
|  | Odonata | 7 genera | 14 species |  |
|  | Plecoptera | *Capnia* | *C. atra* | @brittain1984 |
|  |  | *Capnia* | *C. bifrons* | @elliott1986 |
|  |  | *Mesocapnia* | *M. oenone* | @brittain1984 |
|  |  | *Nemurella* | *N. pictetii* | @brittain1978, @elliott1984 |
|  |  | *Taeniopteryx* | *T. nebulosa* | @brittain1977 |
| Malacostraca | Decapoda | *Pontastacus* | *P. leptodactylus* | @aydin2004 |
| Reptilia | Squamata | *Podarcis* | *P. muralis* | @vandamme1992 |
|  |  | *Sceloporus* | *S. undulatus* | @angilletta2000 |
|  | Testudines | *Mauremys* | *M. reevesii* | @du2007 |
|  |  |  | 181 species | 141 studies in @while2018 |

@@

Figure 1: Custom hatching phenology model for coastal tailed frogs (*Ascaphus truei*). Panel A represents the model fit and raw data used to generate the effective value model and panel B illustrates the effective values for daily temperatures between 6 and 20 °C. A dashed line with a 0.01 increase for every degree increase is included for reference.

Figure 2: Modeled changes in developmental timing for coastal tailed frogs (*Ascaphus truei*) from 1991 to 2020. Panel A: Marginal mean trend across all sites shows a significant decline in days to hatching over time. Panel B: Site-specific (COMID) trends in predicted days to hatch, colored by mountain range. Estimates are based on the best fitting model, with stream temperatures from @siegel2023 and developmental timing modeled using hatchR. Shaded ribbons represent 95% confidence intervals.
